Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2005-04-09 paper/metamonads.tex
change title as per email.  minor edits in intro.

\begin{document}
\title{Functors, CPS and monads, or how to generate efficient
code from abstract designs}
\author{Jacques Carette\inst{1}
 \and
Oleg Kiselyov\inst{2}
}
\institute{McMaster University,
1280 Main St. West, Hamilton, Ontario Canada L8S 4K1
\and
FNMOC, Monterey, CA 93943}

@@

In high-performance, symbolic, and numeric computing, there is a
well-known issue of balancing between maximal performance and the
level of abstraction at which code is written.  Furthermore, already
in linear algebra, there is a wealth of different aspects that
\emph{may} need to be addressed. For example, implementations of the
widely used Gaussian Elimination (GE) algorithm --- the running
example of our paper --- may need to account for the representation of
the matrix, whether to compute and return the detrminant or rank, how
and whether search for pivot, etc. Furthermore, current architectures
demand more and more frequent tweaks which, in general, cannot be done by the
compiler because the tweaking often involves domain knowledge. 
The survey \cite{carette04} of
Gaussian elimination implementation in an industrial package Maple
found XXX various aspects and XXX different implementations of the
algorithm.  We can manually write each of these implementations
optimizing for particular aspects and using cut-and-paste to ``share''
similar pieces of code. \oleg{I don't think we need to say much here
-- the shear number of various implementations should be enough of a
clue.} We can write a very generic GE procedure that accounts for
all the aspects with appropriate abstractions (cite axiom?). The
abstraction mechanisms however -- be they procedure, method or a
function call -- have a significant cost, especially for
high-performance numerical computing (cite? At least carette04).

A more appealing (acceptable? indispensable? need a better word)
approach is generative programming
\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}.
The approach is not without problems, e.g., making sure that the
generated code is well-formed. This is a challenge in string-based
generation systems, which generally do not offer such guarantees and
therefore make it very difficult to determine which part of the
generator is at fault when the generated code cannot be parsed. Other
problems is preventing accidental variable capture (so-called hygiene
\cite{HygienicMacros}) and ensuring the generated code is
well-typed. Lisp-style macros, Scheme hygienic macros, camlp4
preprocessor (cite XXX), C++ template meta-programming, Template
Haskell \cite{conf/dagstuhl/CzarneckiOST03} solve some of the above
problems. Of the widely available maintainable languages, only
MetaOCaml \cite{CTHL03,metaocaml-org}  solves all the above problems
including the well-typing of the generated code \cite{TahaSheard97,TahaThesis}.

@@

In this paper we report on progress of solving these problems using GE
as our running example. Specifically, our contributions:
\vspace*{-6pt}
\begin{itemize}
    \item Extending monad of \cite{KiselyovTaha} for generating
      binding statements when generating control structures such as
      loops and conditionals. We argue that code generation is an
      effect and so requires something like a monad for proper
      serialization.
    \item Implementation of the |mdo|-notation (patterned after
      |do|-notation of Haskell) to make monadic code readable
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{assure composability of aspects} even
      for aspects that use state and have to be accounted in many
      places in the generated code
    \item use functor type sharing constraints to encode domain-specific
      knowledge
\end{itemize}

@@

A more generator |param_code1| has two occurrences of |plus y one|,
which may be quite a complex computation and so we would rather not do
it twice. We may be tempted to rely on the compiler's
common-subexpression elimination optimization. When the generated code is
very complex, however, the compiler may overlook common subexpressions.  Or the
subexpressions may occur in such an imperative context where the compiler
might not be able to determine if lifting them is sound. So, being
conservative, the optimizer will leave the duplicates as they are. We
may attempt to eliminate subexpressions as in |param_code1'|. However,
the result of |param_code1' plus one| still exhibits duplicate
sub-expressions.  Our |let|-insertion optimization saved the
computation at the generating stage.  We need a combinator that
inserts the |let| expression in the generat\emph{ed} code. We need a
combinator |letgen| to be used as\\
|let ce = letgen (plus y one) in plus ce (plus x ce)|
yielding the code like |.<let t = y + 1 in t + (x + t)>.| 
But that seems impossible because |letgen exp| has to generate
the expression |.<let t = exp in body>.| but |letgen| does not
have the |body| yet. The body needs a temporary identifier |.<t>.|
that is supposed to be the result of |letgen| itself.  Certainly
|letgen| cannot generate only part of a let-expression, without the
|body|, as all generated expressions in MetaOCaml are well-formed and
complete.

>>>>

\begin{document}
%\title{Functors, CPS and monads, or how to generate efficient
%code from abstract designs}
\title{Multi-stage programming with functors and monads:
eliminating abstraction overhead from generic code}
\author{Jacques Carette\inst{1}
 \and
Oleg Kiselyov\inst{2}
}
\institute{McMaster University,
1280 Main St. West, Hamilton, Ontario Canada L8S 4K1
\and
FNMOC, Monterey, CA 93943}

@@

In high-performance, symbolic, and numeric computing, there is a
well-known issue of balancing between maximal performance and the
level of abstraction at which code is written.  Furthermore, already
in linear algebra, there is a wealth of different aspects that
\emph{may} need to be addressed. For example, implementations of the
widely used Gaussian Elimination (GE) algorithm --- the running
example of our paper --- may need to account for the representation of
the matrix, whether to compute and return the detrminant or rank, how
and whether search for pivot, etc. Furthermore, current architectures
demand more and more frequent tweaks which, in general, cannot be done by the
compiler because the tweaking often involves domain knowledge. 
A survey \cite{carette04} of
Gaussian elimination implementations in an industrial package Maple
found 6 different aspects and 35 different implementations of the
algorithm, as well as 45 implementations of directly related
algorithms.  We can manually write each of these implementations
optimizing for particular aspects and using cut-and-paste to ``share''
similar pieces of code.
We can write a very generic GE procedure that accounts for
all the aspects with appropriate abstractions \cite{Axiom}. The
abstraction mechanisms however -- be they procedure, method or a
function call -- have a significant cost, especially for
high-performance numerical computing (cite? At least carette04).

A more appealing approach is generative programming
\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}.
The approach is not without problems, e.g., making sure that the
generated code is well-formed. This is a challenge in string-based
generation systems, which generally do not offer such guarantees and
therefore make it very difficult to determine which part of the
generator is at fault when the generated code cannot be parsed. Other
problems is preventing accidental variable capture (so-called hygiene
\cite{HygienicMacros}) and ensuring the generated code is
well-typed. Lisp-style macros, Scheme hygienic macros, camlp4
preprocessor (cite XXX), C++ template meta-programming, Template
Haskell \cite{conf/dagstuhl/CzarneckiOST03} solve some of the above
problems. Of the widely available maintainable languages, only
MetaOCaml \cite{CTHL03,metaocaml-org}  solves all the above problems
including the well-typing of the generated code \cite{TahaSheard97,TahaThesis}.

@@

In this paper we report on progress of solving these problems using GE
as our running example. Specifically, our contributions:
\vspace*{-2pt}
\begin{itemize}
    \item Extending monad of \cite{KiselyovTaha} for generating
      binding statements when generating control structures such as
      loops and conditionals. We argue that code generation is an
      effect and so requires something like a monad for proper
      serialization.
    \item Implementation of the |mdo|-notation (patterned after
      |do|-notation of Haskell) to make monadic code readable
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{assure composability of aspects} even
      for aspects that use state and have to be accounted in many
      places in the generated code
    \item use functor type sharing constraints to encode domain-specific
      knowledge
\end{itemize}

@@

The generator |param_code1| has two occurrences of |plus y one|,
which may be quite a complex computation and so we would rather not do
it twice. We may be tempted to rely on the compiler's
common-subexpression elimination optimization. When the generated code is
very complex, however, the compiler may overlook common subexpressions.  Or the
subexpressions may occur in such an imperative context where the compiler
might not be able to determine if lifting them is sound. So, being
conservative, the optimizer will leave the duplicates as they are. We
may attempt to eliminate subexpressions as in |param_code1'|. However,
the result of |param_code1' plus one| still exhibits duplicate
sub-expressions.  Our |let|-insertion optimization saved the
computation at the generating stage.  We need a combinator that
inserts the |let| expression in the generat\emph{ed} code. We need a
combinator |letgen| to be used as\\
|let ce = letgen (plus y one) in plus ce (plus x ce)|
yielding the code like |.<let t = y + 1 in t + (x + t)>.| 
But that seems impossible because |letgen exp| has to generate
the expression |.<let t = exp in body>.| but |letgen| does not
have the |body| yet. The body needs a temporary identifier |.<t>.|
that is supposed to be the result of |letgen| itself.  Certainly
|letgen| cannot generate only part of a let-expression, without the
|body|, as all generated expressions in MetaOCaml are well-formed and
complete.
